---
title: "exercise-10"
format: html
editor: visual
---

# Exercise 10: ANOVA Practice

```{r}
#Setup
#| warning: false
#| eval: false

library(tidyverse)
library(dplyr)
library(skimr)
d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/AVONETdataset1.csv", col_names = TRUE)

keep <- c("Species1", "Family1", "Order1", "Beak.Length_Culmen", "Beak.Width", "Beak.Depth", "Tarsus.Length", "Wing.Length", "Tail.Length", "Mass", "Habitat", "Migration", "Trophic.Level", "Trophic.Niche", "Min.Latitude", "Max.Latitude", "Centroid.Latitude", "Primary.Lifestyle", "Range.Size")

d <- d |> select(all_of(keep))
```

```{r}
#Data Exploration
skim(d)
```

*There are 7 character variables and 12 numeric, but this doesn't necessarily reflect which are continuous and which are categorical. Character variables include: Species, Family, Order, Habitat, Trophic Level, Trophic.Niche, and Primary.Lifestyle. In addition, Migration is categorical, though it is recorded numerically, which means it needs to be treated as a 'factor' rather than a 'numeric'. The continuous variables include: Beak Length, Beak Width, Beak Depth, Tarsus Length, Wing Length, Tail Length, Mass, Minimum Latitude, Maximum Latitude, Centroid Latitude, and Range Size.*

## Challenge 1

Step 1: Boxplots

*"1" is a code indicating relationship between Trophic Level and log(Mass).*

*"2" is a code indicating relationship between Migration category and log(Mass).*

```{r}
p1 <- ggplot(data = d |> drop_na(Trophic.Level), 
             aes(x = Trophic.Level, y = log(Mass))) +
  geom_boxplot() + 
  geom_jitter(alpha = 0.1)
p1
```

```{r}
p2 <- ggplot(data = d |> drop_na(Migration), 
             aes(x = as.factor(Migration), y = log(Mass))) +
  geom_boxplot() +
  geom_jitter(alpha = 0.1)
p2
```

Step 2: Linear Models

```{r}
m1 <- lm(log(Mass) ~ Trophic.Level, data = d)
summary(m1)
```

There is significant difference between all categories except for Omnivore/Carnivore (p = 0.73).

```{r}
m2 <- lm(log(Mass) ~ as.factor(Migration), data = d)
summary(m2)
```

P values are all significant for m2.

-   Examine the output of the resultant linear models. Is log(Mass) associated with either Trophic.Level or Migration category? That is, in the global test of significance, is the F statistic large enough to reject the null hypothesis of an F value of zero?

-   Given the regression coefficients returned for your **Migration** model, which **Migration** categor(ies) are different than the reference level? What level is the reference level? Relevel and assess differences among the remaining pair of **Migration** categories.

Step 3: Tukey Test

Step 4: Permutation Approach

## Challenge 2

Step 1: New Variables

Step 2: Boxplots

Step 3: ANOVA

Step 4: Passeriformes Only

Step 5: Two-Factor ANOVA

Step 6: with Interaction

Step 7: Visualization

Step 8: Checking ANOVA Requirements
